---
title: "Accessing the Human Cell Atlas datasets"
author:
- name: Federico Marini
  affiliation: 
  - &id1 Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  - Center for Thrombosis and Hemostasis (CTH), Mainz
  email: marinif@uni-mainz.de
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('HCAData')`"
output: 
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Accessing the Human Cell Atlas datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{HCAData}
---

**Compiled date**: `r Sys.Date()`

**Last edited**: 2018-11-05

**License**: `r packageDescription("HCAData")[["License"]]`

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    error = FALSE,
    warning = FALSE,
    message = FALSE
)
```

# Load `HCAData`

The `HCAData` data package provides access to 

Snapshot of date: ...

Pkg content available...

```{r}
message("Doing good!")
```

Accessing individual datasets

```{r}

```

Performing some analysis?

```{r}

```

Or displaying the analysis already done on this?


# Explore data with `iSEE`

The datasets are provided in the form of a `SingleCellExperiment` object. 
A natural companion to this data structure is the `r Biocpkg("iSEE")` package, which can be used for interactive and reproducible data exploration.

```{r, eval=FALSE}
library(iSEE)
iSEE(sce_cordblood)
iSEE(sce_bonemarrow)
```

# Session info

```{r}
sessionInfo()
```

