---
title: "Download and preprocess the Human Cell Atlas bone marrow data (Census of Immune Cells)"
author: "Federico Marini"
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 100)
```

# Helper functions

Following the definitions provided [here](https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/latest/advanced/h5_matrices)...

```{r}
library(HDF5Array)
library(SingleCellExperiment)

getGenomes <- function(path) {
  genomes <- attributes(h5dump(path, load = FALSE))$names
  genomes
}

getRowData <- function(path, genome = NULL) {
  if (is.null(genome)) 
    genome <- getGenomes(path)[1]
  data.frame(
    id = as.character(h5read(path, paste0(genome,"/genes"))),
    symbol=as.character(h5read(path, paste0(genome,"/gene_names"))),
    stringsAsFactors=FALSE
  )
}

getColData <- function(path, genome = NULL) {
  if (is.null(genome)) 
    genome <- getGenomes(path)[1]
  
  barcode <- as.character(h5read(path, paste0(genome,"/barcodes")))
  
  data.frame(
    Barcode=barcode,
    stringsAsFactors=FALSE
  )
}
```

# Downloading the h5 file

(accessed November 6, 2018)

```{r}
url <- "https://s3.amazonaws.com/preview-ica-expression-data/ica_bone_marrow_h5.h5"
path <- basename(url)
download.file(url, path)
```

# Loading the data and preprocessing

Saving `rowData` and `colData`, used when constructing the `SingleCellExperiment` object

```{r}
saveRDS(getRowData(path), "ica_bone_marrow_h5_rowData.rds")
saveRDS(getColData(path), "ica_bone_marrow_h5_colData.rds")
```

Converting into a `HDF5Matrix` object

```{r}
mygenome <- getGenomes("ica_bone_marrow_h5.h5")
tenxmat <- TENxMatrix(path,group = mygenome)
dim(tenxmat)
pryr::object_size(tenxmat)
```

Writing out the object as a rectangular h5 file

```{r}
options(DelayedArray.block.size=1e9) # 1GB block size.
mat.out <- writeHDF5Array(
  tenxmat,
  file="ica_bone_marrow_h5_rectangular.h5",
  name="counts",
  chunkdim=beachmat::getBestChunkDims(dim(tenxmat))
)
```

# Generating and saving the `SingleCellExperiment` object

```{r}
sce_bonemarrow <- 
  SingleCellExperiment(
    list(counts = HDF5Array("ica_bone_marrow_h5_rectangular.h5", "counts"))), 
    rowData = getRowData(path), 
    colData = getColData(path)
  )

sce_bonemarrow

saveRDS(sce_bonemarrow, file = "sce_HCA_bonemarrow.rds")
```

# Session info

```{r}
sessionInfo()
```

